---
title: "Low level 101"
author: "Karla Otoude"
format: html
---
## Vectors, variables, 
###

Rattrape le début !!

A limited number of types that can be used to create complicated objects of different classes : 
- logical 
- factor
- list

```{r}
typeof(1)
typeof(2L)
typeof("alice")
typeof(TRUE)
typeof(iris)
```
### Vector

A vector is a collection of values of the same type (by opposition list is an arbitrary collection of values)

```{r}
y <- c(1, 2, -14, 42)
y
```
Direct access to vector elements is implemented by the bracket operator `[]`. 
Beware that the indexes start at 1. 

```{r}
y[3]
```

```{r}
y[0]
y[-1]
```
Vector of 0 is a numeric because the class of the vector is numeric
vector of -1 removes the first element 

Negative values can be used to "remove" elements 


```{r}
1:6
typeof(1:6)
```
The function seq gives a sequence that starts at the first value, ends at the second and proceed by steps given by the 3rd parameter (here we go 2 by 2)
```{r}
seq(2, 6, 2)
```


### Complex indexing
```{r}
z <- c("Alice", "Bob", "Charlie", "Dylan")
z
z[3:4]
```

```{r}
z
z[2]
z[-1:-3]
```

```{r}
foo <- 10:60
bar <- seq(1, 50, 4)
foo
foo[bar]
```

```{r}
x <- 1:5
y <- c(TRUE, FALSE, FALSE, TRUE, TRUE)
x(y)
```


### Vector oriented calculation

```{r}
x <- 1:5
y <- 6:10
x + y
x - y
x < y
```
Calculations possible because they have the same length

```{r}
x <- 10:20
x - 5
x[x>15]
```
Useful to check various thing about a dataset (compare x under various views)
